### 2020-03-23 Paper_shortened_pdf.Rmd
Address Reviewer 1 comments

\vspace{\baselineskip}
Emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their children). Nevertheless, the existing literature has two major shortcomings. Firstly, in assessing linkages with children’s emotional adjustment, quantitative involvement and qualitative parenting behaviors are rarely jointly investigated. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. Children’s emotional adjustment might, for example, be differently impacted by a parent's warmth if that parent is generally highly involved with the day-to-day minutiae of childrearing, than when (s)he is generally absent. Conversely, children’s emotional adjustment might elicit different qualitative parenting behaviors from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be sanctioned. The present study set out to address this question.

Secondly, although fathers have been theorized to play a unique role in emotional adjustment development, empirical work has rarely addressed the joint and relative contributions of fathers' and mothers' involvement. The present study overcomes these limitations by incorporating both quantitative and quantitative measures of mothers' and fathers' involvement. We use sophisticated methods to examine how joint and relative quantitative involvement moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

## Fatherhood theory

Attachment relationships are grounded in a balance between security and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Fatherhood researchers have posited that warmth and security are more central to mother-child attachment relationships, whereas father-child relationships have been construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. Consequently, fathers and mothers are theorized to influence emotional adjustment development in somewhat different ways.

The past decades have seen major strides in the consolidation of theories to guide fatherhood research. Early fatherhood research primarily operationalized involvement in *quantitative* terms: Presence versus absence in divorced families, and the sharing of childcare duties in dual-earner families [@lambBiosocialPerspectivePaternal1987]. Over time, the notion the fathers should share child-rearing responsibilities with mothers, adopting roles historically fulfilled by mothers, gained ground [@pleckFatherhoodIdealsUnited1997]. Consequently, definitions of father involvement evolved to incorporate *qualitative* dimensions, such as warmth and control [@maccobySocializationContextFamily1983; @pleckPaternalInvolvementRevised2004].

It is now widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004]. Nevertheless, these dimensions are rarely investigated together, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

The present study aims to make several contributions. First, we include both mothers and fathers, in order to tease apart their common and unique roles in children's emotional adjustment development. Second, we incorporate both qualitative and quantitative measures of involvement, and examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, we disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Finally, we examined child sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

We employed data from the `r masked("K-cohort")` of the `r masked("Longitudinal Study of Australian Children (LSAC)")`, waves `r masked("2-5")`, which were collected by means of biannual home visits, and self-report mail questionnaires, from 2006-2012. Each wave was reviewed and approved by the
`r masked("Australian Institute of Family Studies Ethics Committee")`, which is a Human Research
Ethics Committee registered with the `r masked("National Health and Medical Research Council")`. Documentation of the ethics approval is available online at `r masked("https://growingupinaustralia.gov.au/sites/default/files/certification_of_ethics_approval_for_lsac.pdf")`. `r masked("LSAC")` employed a clustered sampling design to obtain a representative sample of the `r masked("Australian")` population. Complete details regarding the study design, sample, and procedure are available online at `r masked("http://data.growingupinaustralia.gov.au/")`.

Our target sample consisted of 1884 native `r masked("Australian")`, non-minority, intact, biological families. At the first wave, children were approximately 6 years old, $M = 6.28$, $SD = .45$, and 12 at the final wave ($M = 12.42$, $SD = .49$); on agerage, mothers were 34.89 ($SD = 4.71$), and fathers 36.94 ($SD = 5.48$). Most families were dual-earner families, based on income from wages or enterprise ($N = 1067$); the remainder were single-earner families($N = 669$ father-only, $N = 41$ mother-only). Income was recorded in ordered categories. The modal annual income for fathers was between \$26,000-\$51,999 ($N = 663$ fathers reported this value), and for mothers, \$25,999 or less per year ($N = 1152$ reported this value).

@@

@@

We calculated mean scores for all of the study variables. Scale descriptive statistics, number of items, number of complete cases, and reliability are reported in Supplementary Table 1, and scale correlations are provided in Supplementary Table 2. We deemed it defensible to assume metric measurement invariance (i.e., to assume that the factor loadings of each construct are the same across time points and respondents), as the difference in fit between structurally invariant models, and metrically invariant models, was smaller than .01 on all conventional fit indices: RMSEA, CFI, and TLI (see Supplementary Table 3).

@@

All analyses were conducted using Mplus 8.0 and the R-package MplusAutomation. This type of multilevel autoregressive model can only be estimated using Bayesian methods. We used two Gibbs-sampler chains of 50000 iterations each. Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing at random. The missing data handling is likelihood based and guarantees consistent estimation as long as the missing data is missing at random (MAR). We applied mildly informative inverse Wishart priors to the variance/covariance matrices, and normal priors centered on zero for all regression coefficients. We assessed convergence through visual inspections of the chains and posterior distributions. We report the posterior median, and a 95% highest posterior density interval (HPDI). This is the "narrowest" possible interval that contains 95% of samples from the posterior distribution. It represents the most likely population values of the parameter in question. Results are best interpreted by examining the effect size (standardized regression coefficients and correlations), and the width of the HPDI. Significance is not a Bayesian concept, but we use the shorthand "significant" to refer to estimates for which less than 5% of the posterior distribution had the opposite sign from the estimate. All standardized and unstandardized model parameters are reported in Supplementary Table 4.

@@

tab$Parameter <- gsub("^Variance", "Unstandardized variance", tab$Parameter)
names(tab)[1] <- "Param."
reg_table(tab, caption = "Standardized between-family level intercepts and unstandardized variances of within-family predictive effects", note = "Param.: Parameter. Est.: Estimate. HDPI: Highest posterior density interval. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate. Matching letters indicate that parameters differed significantly; see Supplementary Table S5 for all difference tests.")
```

With regard to child effects, in line with hypothesis $H_{2a}$, the intercept of the cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more warmth than usual in the next wave. Contrary to hypothesis $H_{2b}$, the intercept of the effect of emotional adjustment on maternal consistency was non-significant. As hypothesized in $H_{2c}$, the intercept for maternal inductive reasoning was positive. Counter to hypothesis $H_{2d}$, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more anger than usual in the next wave. In support of hypothesis $H_{2a}$, the intercept for the child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts ($H_{2b-d}$)for child effects on paternal parenting were significant. The variances of all child effects were all significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

We established direction of effects by testing the significance of the difference between corresponding parent- and child effects (see Table \@ref(tab:tabcrosspath) and Supplementary Table S5). All effects that were significant in one direction, and non-significant in the opposite direction, had significant difference tests. All effects that were significant (or non-significant) in both directions had non-significant difference tests. We examined differences between fathers and mothers in parent- and child effects the same way. All effects that were significant for one parent but not significant for the other parent had significant difference tests, except for the child effect of emotional adjustment on parental anger, which did not differ significantly between parents.

@@

We examined standardized regression coefficients of joint and relative quantitative involvement on the random intercepts of qualitative parenting behaviors and emotional adjustment (Table \@ref(tab:tabinvolvement)). With regard to parenting, we found partial support for our hypotheses $H_{4a-c}$: For both parents, joint involvement positively predicted greater warmth, consistency, and inductive reasoning. Notably, the association between joint involvement and warmth was almost double in size for fathers, as compared to mothers. We did not find the hypothesized negative effects of joint involvement on parental anger, $H_{4d}$. With regard to relative involvement, we found limited support for hypotheses $H_{5a-d}$, that greater differential involvement would attenuate the desirable parenting behaviors of the more involved parent positively, and those of the less involved parent negatively. For mothers, the only significant association we found was that mothers displayed more warmth when they were relatively more involved than fathers. We found that fathers displayed less warmth, consistency, and inductive reasoning, when mothers were relatively more involved than fathers.

Counter to our hypothesis $H_{6a}$, we found that greater joint involvement was negatively associated with emotional adjustment. Thus, the children in families with highly involved parents generally displayed poorer emotional adjustment. However, hypothesis $H_{6b}$, that children's emotional adjustment would be lower in families where mothers were relatively more involved than fathers, was supported.

@@

We found sex differences in the random intercepts of qualitative parenting behaviors: In partial support of $H_{7a,d}$, fathers displayed more warmth towards daughters, and both parents displayed more anger towards sons (Table \@ref(tab:tabsex)). Counter to hypothesis $H_{7c}$, both parents also displayed more inductive reasoning towards sons. We found no sex differences in the random intercept of emotional adjustment. Hypotheses $H_{8a-d}$ were rejected: There was no evidence for stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent. The only predictive effects moderated by child sex were child effects of emotional adjustment on maternal and paternal inductive reasoning. Emotional adjustment was more negatively predictive of parents' inductive reasoning for boys, as compared to girls. In other words, if children display worse emotional adjustment than usual in one wave, parents increase inductive reasoning more strongly in the next wave for boys than for girls.

@@

In terms of parenting effects, the only effect that was significant on average was that maternal warmth positively predicted emotional adjustment in most families. However, all within-family predictive effects significantly varied across families, and this variability was notably greater for fathers, indicating that fathers' influence differs most across families. Our results thus emphasize the fact that dynamic reciprocal processes vary from one family to the next.

Regarding child effects, we found that, on average, well-adjusted children evoked greater warmth from both parents, as hypothesized. Counter to predictions, we did not find average child effects for consistency. In partial support of our hypothesis, we found that children who displayed better emotional adjustment in one wave evoked greater inductive reasoning from mothers in the next wave. But paradoxically, and counter to oru hypothesis, we also found that children who displayed better emotional adjustment in one wave evoked greater maternal anger in the next wave.

@@

For anger, we found moderately strong between-family level correlations for both parents. Counter to our expectations, however, we found no evidence for within-family correlations or parenting effects. This suggests that the brunt of negative associations between anger and adjustment is explained by between-family differences. One potential explanation for the absence of within-family associations might be a ceiling effect: If the stable level of parental anger in a family is high enough to negatively impact children's emotional adjustment, over-time fluctuations of that anger might no longer make much of a difference. We also found a paradoxical positive predictive effect of emotional adjustment on maternal anger. This means that, if children demonstrated better emotional adjustment than usual in one wave, their mothers tended to display more anger than usual in the next wave. A possible explanation might be regression to the mean: On average, emotional adjustment showed no-to-negative stability coefficients in all models. That means that better emotional adjustment in one wave, was likely to be followed worse adjustment in the next wave. The literature on expectation-based loss aversion suggests that mothers might be more angered by a child who unexpectedly displays poor adjustment, than they would be appeased by a child who displays unexpected positive adjustment [@koszegiModelReferenceDependentPreferences].

@@

<!-- To do!! --> 
Overall, we found somewhat more evidence for associations of emotional adjustment with mothers' qualitative parenting, than with fathers' parenting, and we found some evidence for differences between mothers' and fathers' roles in emotional adjustment development. These differences all played out at the within-family level, however. Stable between-family differences in either parent's qualitative parenting were all similarly associated with children's adjustment.

At the within-family level we found stronger associations with emotional adjustment for maternal consistency, and for paternal reasoning and warmth. As mentioned before, mothers tend to spend more time with their children than fathers; also in our sample. This might help explain the stronger associations for maternal consistency - a behavior which implies, by definition, consistent repeated interactions with the child. For fathers, associations were stronger for inductive reasoning and warmth - behaviors which can also be expressed in one-off interactions, such as when  disciplining a child (inductive reasoning), or during play time (warmth). The stronger association between inductive reasoning and emotional adjustment for fathers is in line with the classic view of the father as disciplinarian [@flouriFathersBehaviorsChildren2010].

@@

We found limited evidence that quantitative involvement moderated within-family parenting effects. One such effect is that the effect of maternal warmth on adjustment was contingent on both joint- and relative quantitative involvement. Mothers' warmth predicted emotional adjustment less strongly when both parents were more involved, and more strongly when mothers were relatively more involved than fathers. These findings suggests a compensatory effect, whereby mothers' warmth can make up for disadvantages incurred by having a relatively uninvolved father [@shinnFatherAbsenceChildren1978].

@@

With regard to mothers' versus fathers' unique roles in children's emotional development, we found that, in line with the notion of the mother as a warm, supportive attachment figure, maternal warmth and consistency were associated with children's emotional adjustment. Evidence for fathers' unique role was relatively limited; significant within-family correlations between paternal inductive reasoning and emotional adjustment suggest that fathers might play a role in the capacity of disciplinarian. The aforementioned interaction effects between qualitative and quantitative involvement, however, seem to suggest that - above all - the role of fathers is more context dependent than that of mothers.

>>>>

\vspace{\baselineskip}
Children's emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their child). Nevertheless, the existing literature has two major shortcomings. Firstly, although it is widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. For example, children's emotional adjustment might be differently impacted by anger coming from a highly involved parent, than from a generally absent parent. Conversely, children's poor emotional adjustment might elicit warm responses from hands-on parents, who are aware of the reasons for their children's emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be disciplined with anger or inductive reasoning. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

Secondly, although emotional adjustment is one particular area of development where fathers have been theorized to play a unique role [@paquetteTheorizingFatherchildRelationship2004], fathers are still not consistently included in research [@faganShouldResearchersConceptualize2014]. Consequently, prior research has not sufficiently addressed mothers' and fathers' joint and unique contributions to emotional adjustment development. The present study overcomes this limitation by incorporating both parents. We use sophisticated methods to examine how joint and relative quantitative involvement at age 6 moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

@@

The present study examined whether specific qualitative parenting behaviors are associated with children's emotional adjustment, and whether these associations are moderated by quantitative involvement. We aimed to make several contributions. First, by including both mothers and fathers, we can tease apart their common and unique roles in children's emotional adjustment development. Second, by incorporating both qualitative and quantitative measures of involvement, we can examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, by disentangling stable between-family differences from dynamic within-family predictive processes, we can more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Finally, we examined child sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

We employed data from the `r masked("K-cohort")` of the `r masked("Longitudinal Study of Australian Children (LSAC)")`, waves `r masked("2-5")`, which were collected by means of biannual home visits, and self-report mail questionnaires, from 2006-2012. Each wave was reviewed and approved by the
`r masked("Australian Institute of Family Studies Ethics Committee")`, which is a Human Research
Ethics Committee registered with the `r masked("National Health and Medical Research Council")`. Documentation of the ethics approval is available online [@australianinstituteoffamilystudiesGrowingAustralia2020]. `r masked("LSAC")` employed a clustered sampling design to obtain a representative sample of the `r masked("Australian")` population. Complete details regarding the study design, sample, and procedure are available online [@australianinstituteoffamilystudiesGrowingAustralia2020].

From the total sample (4.983 focal children at the first wave), we selected the 1884 native Australian, non-minority, intact biological families that constituted our target population. Although minority- or immigration status and family composition are important factors, they are beyond the scope of this paper. At the first wave, children in our sample were approximately 6 years old, $M = 6.28$, $SD = .45$, and 12 at the final wave ($M = 12.42$, $SD = .49$); on average, mothers were 34.89 ($SD = 4.71$), and fathers 36.94 ($SD = 5.48$). Most families were dual-earner families, based on income from wages or enterprise ($N = 1067$); the remainder were single-earner families($N = 669$ father-only, $N = 41$ mother-only). The LSAC did not measure income as a continuous variable, but as an ordered categorical variable. For fathers, the modal category was "\$26,000-\$51,999" ($N = 663$ fathers reported this value), and for mothers, "\$25,999 or less per year" ($N = 1152$ reported this value).

@@

We calculated mean scores for all study variables with multiple indicators. Scale descriptive statistics, number of items, attrition (indicated by the number of complete cases), and reliability are reported in Supplementary Table 1, and scale correlations are provided in Supplementary Table 2. The difference in fit between structurally invariant models and metrically invariant models was smaller than .01 on all conventional fit indices: RMSEA, CFI, and TLI (see Supplementary Table 3). We thus deemed it defensible to assume metric measurement invariance (i.e., to assume that the factor loadings of each construct are the same across time points and respondents).

@@

@@

Analyses were conducted using Mplus 8.0 and the R-package MplusAutomation. This type of multilevel autoregressive model can only be estimated using Bayesian methods. We used two Gibbs-sampler chains of 50000 iterations each. Bayes is a full-information estimator, which means that all available information is used without imputing missing values, and consistent estimation is guaranteed under the assumption that data are missing-at-random (a non-parametric MCAR test resulted in $p > .05$). We used mildly informative inverse Wishart priors for covariance matrices, and normal priors centered on zero for regression coefficients. We assessed convergence through visual inspection of the chains and posterior distributions. We report the posterior median, and a 95% highest posterior density interval (HPDI). This is the "narrowest" possible interval that contains 95% of samples from the posterior distribution. It represents the most likely population values of the parameter in question. Results are best interpreted by examining the effect size (correlations and standardized regression coefficients), and the width of the HPDI. Significance is not a Bayesian concept, but we use the shorthand "significant" to refer to estimates for which less than 5% of the posterior distribution had the opposite sign from the estimate. Supplementary Table S4 reports all standardized and unstandardized model parameters. All code and output are available online at [masked].

@@

tab$Parameter <- gsub("^Variance", "Unstandardized variance", tab$Parameter)
names(tab)[1] <- "Param."
reg_table(tab, caption = "Standardized between-family level intercepts and unstandardized variances of within-family predictive effects", note = "Param.: Parameter. Est.: Estimate. HDPI: Highest posterior density interval. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate. Matching letters indicate that parameters differed significantly; see Supplementary Table S4 for all difference tests.")
```

With regard to child effects, in line with hypothesis $H_{2a}$, the intercept of the cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more warmth than usual in the next wave. Contrary to hypothesis $H_{2b}$, the intercept of the effect of emotional adjustment on maternal consistency was non-significant. As hypothesized in $H_{2c}$, the intercept for maternal inductive reasoning was positive. Counter to hypothesis $H_{2d}$, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more anger than usual in the next wave. In support of hypothesis $H_{2a}$, the intercept for the child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts ($H_{2b-d}$) for child effects on paternal parenting were significant. The variances of all child effects were all significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

We established direction of effects by testing the difference between corresponding parent- and child effects (see Table \@ref(tab:tabcrosspath) and Supplementary Table S4). All effects that were significant in one direction, and non-significant in the opposite direction, had significant difference tests. All effects that were significant (or non-significant) in both directions had non-significant difference tests. We examined differences between fathers and mothers in parent- and child effects the same way. All effects that were significant for one parent but not significant for the other parent had significant difference tests, except for the child effect of emotional adjustment on parental anger, which did not differ significantly between parents.

@@

We examined standardized regression coefficients of joint and relative quantitative involvement on the random intercepts of qualitative parenting behaviors and emotional adjustment (Table \@ref(tab:tabinvolvement)). With regard to parenting, we found partial support for our hypotheses $H_{4a-c}$: For both parents, joint involvement positively predicted greater warmth, consistency, and inductive reasoning. Notably, the association between joint involvement and warmth was almost double in size for fathers, as compared to mothers. We did not find the hypothesized negative effects of joint involvement on parental anger, $H_{4d}$. With regard to relative involvement, we found limited support for hypotheses $H_{5a-d}$, that greater differential involvement would attenuate the desirable parenting behaviors of the more involved parent positively, and those of the less involved parent negatively. For mothers, the only significant association we found was that mothers displayed more warmth when they were relatively more involved than fathers. We found that fathers displayed less warmth, consistency, and inductive reasoning, when mothers were relatively more involved than fathers. Counter to hypothesis $H_{6a}$, we found that greater joint involvement negatively predicted emotional adjustment. Thus, children in families with highly involved parents generally displayed poorer emotional adjustment. Hypothesis $H_{6b}$ was supported: Children's emotional adjustment was lower in families where mothers were relatively more involved than fathers.

@@

We found sex differences in the random intercepts of qualitative parenting behaviors: In partial support of $H_{7a,d}$, fathers displayed more warmth towards daughters, and both parents displayed more anger towards sons (Table \@ref(tab:tabsex)). Counter to hypotheses $H_{7b,c}$, there were no sex differences in consistency, and both parents displayed more inductive reasoning towards sons. We found no sex differences in the random intercept of emotional adjustment. Hypotheses $H_{8a-d}$ were rejected: There was no evidence for stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent. The only predictive effects moderated by child sex were child effects of emotional adjustment on maternal and paternal inductive reasoning. Emotional adjustment was more negatively predictive of parents' inductive reasoning for boys, as compared to girls. In other words, if children display worse emotional adjustment than usual in one wave, parents increase inductive reasoning more strongly in the next wave for boys than for girls.

@@

The only parenting effect that was significant on average, was for maternal warmth, which positively predicted emotional adjustment in most families. All within-family predictive effects varied significantly across families, however, which highlights the importance of capturing and explaining these differences. This variability was notably greater for fathers than mothers, which suggests that fathers' influence is more context-dependent.

Regarding child effects, as hypothesized, well-adjusted children evoked greater warmth from both parents on average. Counter to predictions, we did not find child effects for consistency. In partial support of our hypothesis, we found that children who displayed better emotional adjustment in one wave evoked greater inductive reasoning from mothers in the next wave. But paradoxically, and counter to our hypothesis, we also found that children who displayed better emotional adjustment in one wave evoked greater maternal anger in the next wave. Perhaps the two go hand in hand: When mothers get angry, they (eventually) also explain what provoked their anger, through inductive reasoning.

@@

For anger, we found moderately strong between-family level correlations for both parents. Counter to our expectations, however, we found no evidence for within-family correlations or parenting effects. This suggests that the brunt of negative associations between anger and adjustment is explained by between-family differences. One potential explanation for the absence of within-family associations might be a ceiling effect: If the stable level of parental anger in a family is high enough to negatively impact children's emotional adjustment, over-time fluctuations of that anger might no longer make much of a difference. We also found a paradoxical positive predictive effect of emotional adjustment on maternal anger. This means that, if children demonstrated better emotional adjustment than usual in one wave, their mothers tended to display more anger than usual in the next wave. A possible explanation might be regression to the mean: On average, emotional adjustment showed no-to-negative stability coefficients in all models. That means that better emotional adjustment in one wave, was likely to be followed worse adjustment in the next wave. The literature on expectation-based loss aversion suggests that mothers might be more angered by a child who unexpectedly displays poor adjustment, than they would be appeased by a child who displays unexpected positive adjustment [@koszegiModelReferenceDependentPreferences2006].

@@

Overall, we found somewhat more evidence for associations of emotional adjustment with mothers' qualitative parenting, than with fathers' parenting, and we found some evidence for differences between mothers' and fathers' roles in emotional adjustment development. These differences all played out at the within-family level, however. Stable between-family differences in either parent's qualitative parenting were all similarly associated with children's adjustment.

At the within-family level, associations with emotional adjustment were stronger for maternal consistency, and for paternal reasoning and warmth. As mentioned before, mothers tend to spend more time with their children than fathers; also in our sample. This might help explain the stronger associations for maternal consistency - a behavior which implies, by definition, consistent repeated interactions with the child. For fathers, associations were stronger for inductive reasoning and warmth - behaviors which can also be expressed in one-off interactions, such as when  disciplining a child (inductive reasoning), or during play time (warmth). The stronger association with inductive reasoning for fathers is in line with the classic view of the father as disciplinarian [@flouriFathersBehaviorsChildren2010].

@@

We found limited evidence that quantitative involvement moderates within-family parenting effects. We did find both joint- and relative quantitative involvement moderated the effect of maternal warmth on adjustment. Mothers' warmth predicted emotional adjustment less strongly when both parents were more involved, and more strongly when mothers were relatively more involved than fathers. These findings suggests a compensatory effect, whereby mothers' warmth can make up for disadvantages incurred by having a relatively uninvolved father [@shinnFatherAbsenceChildren1978].

@@

With regard to mothers' versus fathers' unique roles in children's emotional development, we found that, in line with the notion of the mother as a warm, supportive attachment figure, maternal warmth and consistency were associated with children's emotional adjustment. Evidence for a consistent unique role for fathers was limited, although within-family correlations between paternal inductive reasoning and emotional adjustment suggest that fathers might play a role as disciplinarian. Effects of fathers' parenting varied significantly more than mothers', however. Combined with the aforementioned interaction effects between qualitative and quantitative involvement, this suggests that - above all - the role of fathers is more context dependent than that of mothers.
